---
title: "wbplot: the World Bank ggplot2 theme"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{wbplot: the World Bank ggplot2 theme}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{=html}
<style>
    img {
        border: 0;
    }
</style>
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What is wbplot?

wbplot is an R package that provides functions for theming ggplot2 visualizations according to the [World Bank Data Visualization Style Guide](https://wbg-vis-design.vercel.app/).

## Installation and loading

Install the package with `devtools::install_github("worldbank/wbplot")` or `remotes::install_github("worldbank/wbplot")`.

You might run into an authentication error. To overcome this, generate a token on [github.com/settings/personal-access-tokens](https://github.com/settings/personal-access-tokens) and run

``` r
devtools::install_github("worldbank/wbplot", auth_token = "MyPersonalToken")
```
When the package is installed, load it into your R session with `library(wbplot)`. All `wbplot` extend the functionality of the `ggplot2` package, and for data manipulation, we use the `dplyr` function.

```{r setup}
library(dplyr)
library(ggplot2)
library(wbplot)
```

## Included data

The package comes with 2 data sets:

-   `countries`: available countries and regions from the World Bank API, as returned by `wbstats::wbcountries`
-   `life.expectancy`: time series data for the SP.DYN.LE00.IN indicator for all countries and regions, as returned by `wbstats::wb_data("SP.DYN.LE00.IN")`

```{r}
print(countries)
```

```{r}
print(life.expectancy)
```

## Styling a bar chart

We can make a standard ggplot bar chart from the life expectancy data for 2022.

```{r, fig.dim = c(8,4), dpi = 300, out.width = "100%"}
country.names <- select(countries, iso3c, country)

life.22 <- filter(life.expectancy, date == 2022) |>
  select(iso3c, SP.DYN.LE00.IN)

life.22.top10 <- head(dplyr::arrange(life.22, desc(SP.DYN.LE00.IN)),10) |>
  left_join(country.names, by = "iso3c")

barchart <- ggplot(life.22.top10, aes(
    x = SP.DYN.LE00.IN,
    y = reorder(country, SP.DYN.LE00.IN)
  )) +
  geom_bar(stat="identity", width = 0.66) +
  geom_text(aes(label = round(SP.DYN.LE00.IN, 1)), nudge_x = 0.7) +
  labs(
    title = "The countries with the longest living people",
    subtitle = "Life expectancy at birth (total (years), 2022)")
barchart
```

Now we can apply the `theme_wb` theme to the plot. This will change the used fonts, the color of the bars, the positioning of the the title and subtitle, and the styling of the grid lines.

```{r, fig.dim = c(8,4), dpi = 300, out.width = "100%"}
barchart + theme_wb()
```

The `wb_theme()` can be added to any ggplot visualization. But the World Bank data visualization style calls for some bar chart specific styling options. We can apply those by setting by setting the `chartType` parameter to "bar" in the theme function:

```{r, fig.dim = c(8,4), dpi = 300, out.width = "100%"}
barchart + theme_wb(chartType = "bar")
```

This will remove the grid lines, remove the axis titles, move the x axis to the top and align the category labels to the left.

It will also avoid scales to expand beyond the values in the data, which might clip some elements on the chart, like data labels added with `geom_text`. You can add some expansion to the x axis to avoid this:

```{r, fig.dim = c(8,4), dpi = 300, out.width = "100%"}
barchart + theme_wb(chartType = "bar", xExpansion = 3)
```

If you need a title for the x axis, you can add it again with the `addXAxisTitle` parameter (which you then should set with `ggplot2::xlab()`).

```{r, fig.dim = c(8,4), dpi = 300, out.width = "100%"}
barchart + theme_wb(chartType = "bar", xExpansion = 3, addXAxisTitle = TRUE)
```

A common element on World Bank data visualizations are notes or source references. You can add a styled note to the bottom left corner of a plot with `add_note_wb()`.

```{r, fig.dim = c(8,5), dpi = 300, out.width = "100%"}
barchart +
  theme_wb(chartType = "bar", xExpansion = 3, addXAxisTitle = FALSE) +
  add_note_wb(noteTitle = "Source:", note = "World Bank")
```

## Styling a scatter plot

Now let's do the same thing for a scatter plot. We can get some additional data on country GDP using with the `wb_data()` function of the `wbstats` package.

```{r}
gdp <- wbstats::wb_data("NY.GDP.PCAP.PP.CD")
gdp.22 <- filter(gdp, date == 2022) |>
  select(iso3c, NY.GDP.PCAP.PP.CD)

scatterdata <- left_join(gdp.22, life.22, by = "iso3c") |>
  left_join(countries, by = "iso3c")
print(scatterdata)
```
With this data, we can create a default ggplot scatterplot.

```{r, fig.dim = c(8,6), dpi = 300, out.width = "100%", warning = FALSE}
scatterplot <- ggplot(scatterdata, aes(NY.GDP.PCAP.PP.CD, SP.DYN.LE00.IN, fill = region_iso3c)) +
  geom_point() +
  labs(
    title = "Richer countries have higher life expectancy",
    subtitle = "GDP (per capita $ PPP) vs total life expectancy at birth (years)",
    x = "GDP per capita",
    y = "Life expectancy"
  )
scatterplot
```
Notice that because you ran the `theme_wb()` function before, the default shape for `geom_point()` was changed: by default it uses a filled circle with a white outline (`geom_point(shape = 21, color = "white")`). That is why we use a mapping to the fill aesthetic instead of the color aesthetic.

Applying `theme_wb` to this plot gives you this:

```{r, fig.dim = c(8,6), dpi = 300, out.width = "100%", warning = FALSE}
scatterplot + theme_wb()
```

Like for bar charts, the World Bank data visualization style has some specific styling for scatter plots:

```{r, fig.dim = c(8,6), dpi = 300, out.width = "100%", warning = FALSE}
scatterplot + theme_wb(chartType = "scatter")
```

Whenever a continuous scale is present, you can include a line to indicate the zero value on the axis. When zero was not included on the axis initially, the scale will extend to include it.

```{r, fig.dim = c(8,6), dpi = 300, out.width = "100%", warning = FALSE}
scatterplot + theme_wb(chartType = "scatter", addXZeroLine = TRUE, addYZeroLine = TRUE)
```

The wbplot package includes continuous and discrete color scales for the World Bank data visualization color palettes. In this case we should use the `scale_fill_wb_d` to apply the default discrete color palette to the regions.

```{r, fig.dim = c(8,6), dpi = 300, out.width = "100%", warning = FALSE}
scatterplot +
  theme_wb(chartType = "scatter", addXZeroLine = TRUE, addYZeroLine = TRUE) +
  scale_fill_wb_d(name = "Region")
```

Each World Bank also has a color assigned to it in the `region` palette, so we can map the region code directly to the corresponding color.

```{r, fig.dim = c(8,6), dpi = 300, out.width = "100%", warning = FALSE}
scatterplot +
  aes(fill = region_iso3c) +
  theme_wb(chartType = "scatter", addXZeroLine = TRUE) +
  scale_fill_wb_d(palette = "region", name = "Region")
```

Income levels also have their own palette, called "income".

```{r, fig.dim = c(8,6), dpi = 300, out.width = "100%", warning = FALSE}
scatterplot +
  aes(fill = income_level_iso3c) +
  theme_wb(chartType = "scatter", addXZeroLine = TRUE) +
  scale_fill_wb_d(palette = "income", name = "Income level")
```

Alternatively, we can map the fill aesthetic to a numeric variable, and use the continuous color scale to apply a World Bank continuous color palette.

```{r, fig.dim = c(8,6), dpi = 300, out.width = "100%", warning = FALSE}
scatterplot +
  aes(fill = NY.GDP.PCAP.PP.CD) +
  theme_wb(chartType = "scatter", addXZeroLine = TRUE) +
  scale_fill_wb_c(palette = "seq", name = "GDP/capita")
```

Binned color scales are also provided:
```{r, fig.dim = c(8,6), dpi = 300, out.width = "100%", warning = FALSE}
scatterplot +
  aes(fill = NY.GDP.PCAP.PP.CD) +
  theme_wb(chartType = "scatter", addXZeroLine = TRUE) +
  scale_fill_wb_b(palette = "divPosNeg", name = "GDP/capita", n.breaks = 8)
```

## Styling a line chart

Let's prepare some time series data and make a ggplot line chart.

```{r, fig.dim = c(8,6), dpi = 300, out.width = "100%", warning = FALSE}
life.series <- filter(life.expectancy, iso3c %in% c("USA", "CHN", "IND", "DEU", "RUS", "IDN", "JPN"))

linechart <- ggplot(life.series, aes(x = date, y = SP.DYN.LE00.IN, color = iso3c)) +
  geom_line(lineend = "round") +
  ggtitle("Life expectancy of selected countries", subtitle = "Life expectancy at birth, total (years)") +
  theme(legend.title = element_blank())
linechart
```

Applying the general `wb_theme` and the line chart specific styling to it, together with the World Bank discrete color scale:

```{r, fig.dim = c(8,6), dpi = 300, out.width = "100%", warning = FALSE}
linechart +
  theme_wb(chartType = "line", addYZeroLine = TRUE) +
  scale_color_wb_d() +
  theme(legend.title = element_blank())
```

We can give more context by providing the data for all countries in the background.

```{r, fig.dim = c(8,6), dpi = 300, out.width = "100%", warning = FALSE}
linechart +
  theme_wb(chartType = "line", addYZeroLine = TRUE) +
  scale_color_wb_d() +
  theme(legend.title = element_blank()) +
  geom_line(
    data = life.expectancy,
    aes(group = iso3c),
    lineend = "round",
    color = WBCOLORS$grey500,
    alpha = 0.15,
    linewidth = 0.25
  ) +
  geom_line(lineend = "round")
```

Here you can see that all colors of the World Bank data visualization style are available through `WBCOLORS`.

## Styling a beeswarm plot

Finally, let's create a beeswarm plot and style it. To plot a beeswarm plot, we use the ggbeeswarm package.

```{r, fig.dim = c(8,6), dpi = 300, out.width = "100%", warning = FALSE}
lifeexp.22 <- filter(life.expectancy, date == 2022) |>
  left_join(countries, by = "iso3c") |>
  filter(income_level_iso3c != "INX") |>
  mutate(income_level_iso3c = factor(income_level_iso3c, levels = c("HIC", "UMC", "LMC", "LIC")))

beeswarm <- ggplot(lifeexp.22, ggplot2::aes(x = SP.DYN.LE00.IN, y = income_level_iso3c, fill = income_level_iso3c)) +
  ggbeeswarm::geom_beeswarm(
    cex = 2,
    method = "swarm",
    priority = "random",
    size = 2.5
  ) +
  ggtitle("Countries with higher income have longer life expectancy", subtitle = "Life expectancy at birth (total (years), 2022)") +
  theme(legend.position = "none")
beeswarm
```

```{r, fig.dim = c(8,6), dpi = 300, out.width = "100%", warning = FALSE}
beeswarm +
  theme_wb(chartType = "beeswarm", addYAxisTitle = TRUE) +
  scale_fill_wb_d(palette = "income") +
  theme(legend.position = "none")
```

## Colors

All colors of the World Bank Data Visualization Style Guide are available through `WBCOLORS`.

```{r}
WBCOLORS$female

WBCOLORS$noData
```

Likewise, all color palettes are available through `WBPALETTES`.

The sequential palettes are `WBPALETTES$seq`, `WBPALETTES$seqRev`, `WBPALETTES$seqB`
, `WBPALETTES$seqY` and `WBPALETTES$seqP`. The diverging palettes are `WBPALETTES$divPosNeg` and `WBPALETTES$divLR`. All these palettes can be used with the continuous and binned color scales (`scale_fill_wb_c()`, `scale_color_wb_c()`, `scale_fill_wb_b()`, `scale_color_wb_b()`), with the palette parameter (eg. `scale_fill_wb_c(palette = "seqY")`).

The discrete color palettes are `default`, `defaultText`, `region`, `regionText`, `income`, `gender`, `urbanisation`, `age` and `binary`. These can be used in `scale_fill_wb_d()` and `scale_color_wb_d()`, with the `palette` parameter. When the values in the mapped column correspond to the names of the colors of the palette, values are automatically matched to the correct color.

```{r}
WBPALETTES$income

WBPALETTES$gender
```

## Saving plots

As always , you can save plots with `ggsave()`. You should be able to get good results by setting `units` to "px" and `width` to 960. The optimal `height` is determined by the chart type and the data, but the default is 540.

```{r, eval = FALSE}
ggsave(filename = "my-worldbank-chart.png", units = "px", width = 960, height = 960, dpi = 200)
```

For convenience, you can use `ggsave_wb(filename = "my-worldbank-chart.png")`, which has these values for units, dimensions and dpi by default.
